---
summary: |
  This tutorial will introduce you to FlyBase ([flybase.org](http://flybase.org)).  FlyBase (short for fly database) is a website for the *Drosophila* scientific community that organizes research results about *Drosophila*.
author:
  - "[Katherine Cox](https://c-moor.github.io/portfolio/coxkatherine/)"
time: "20 min"
prerequisites: "None"
---

<!-- I included these questions in reflecton boxes, as short answer rather than multiple choice -->

```{r, flybase-lab, echo=FALSE}
quiz(caption = "Look up the *lab* gene on FlyBase",
  question("What is the FlyBase ID number for *lab*?",
    answer("FBgn0002522", correct=TRUE),
    answer("FBgn0000024"),  
    answer("FBgn0000003"),
    answer("FBgn0001264"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What biological processes is the *lab* gene involved in? HINT: Choose more than one answer.",
    answer("brain development", correct=TRUE),
    answer("midgut development", correct=TRUE),
    answer("immune system"),
    answer("cell cycle"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

## Finding Orthologs

```{r huntingtons-lab-orthologs}
quiz(caption = "Investigate orthologs for the *lab* gene",
  question("What is the most likely human ortholog for the *lab* gene?",
    answer("HTT"),
    answer("HOXA", correct = TRUE),
    answer("HSAP"),
    answer("LAT"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("How many programs think this gene is an ortholog of *lab*?",
    answer("1"),
    answer("4"),
    answer("7", correct = TRUE),
    answer("11"),
    answer("15"),
    allow_retry = TRUE
  )
)
```

## Disease models

```{r huntingtons-quiz-1}
quiz(caption = "Explore the Huntington's disease report",
  question("What is the gene symbol for the Drosophila ortholog of the human Huntington's disease gene?",
    answer("hunt"),
    answer("HD"),
    answer("htt", correct = TRUE),
    answer("Dmel"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

### Investigate the *htt* gene

Now that we know the *htt* gene is used to model Huntington’s disease, let’s find out more about this gene.  Look up the Gene report for this gene.  The easiest way to do this is to type "htt" in the Jump to Gene box.

```{r huntingtons-quiz-2}
quiz(caption = "Explore the *htt* gene report",
  question("What biological processes is the *htt* gene involved in? HINT: Choose more than one answer.",
    answer("brain development"),
    answer("choline catabolic process"),
    answer("establishment of mitotic spindle orientation", correct = TRUE),
    answer("synaptic vesicle transport", correct = TRUE),
    answer("cell cycle"),
    answer("axo-dendritic transport", correct = TRUE),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("How many different *alleles* (versions of the gene) can be used as models of Huntington's disease, **based on experimental evidence**?",
    answer("0"),
    answer("1"),
    answer("2"),
    answer("3", correct = TRUE),
    answer("4"),
    allow_retry = TRUE
  )
)
```

### *htt* alleles (variant forms of the gene)

Yes!  Here we can see that there are actually three different variants of the *htt* gene that have been experimentally shown to cause symptoms similar to Huntington’s disease in flies.  These variants all have different types of mutations affecting the *htt* gene.

It can be useful to have different versions of the gene that all mimic Huntington's disease.  Depending how these mutations were made or discovered, they can be useful for different types of experiments.  Also, if experiments turn out the same for several different disease-causing mutations, it's more likely that the result will transfer to other organisms, and isn't just a quirk of a particular mutation.

## Summary

FlyBase is a fantastic source of information.  You can use FlyBase to find:

- Gene functions and interactions
- Orthologs (*Drosophila* genes that are related to genes in other species, including humans)
- *Drosophila* genes that are relevant to human diseases
- The name, symbol or ID number for a gene, so you can research it on other databases.

#### Learn More

FlyBase is designed to help scientists do research, so it has *a lot* of information, and we've just looked at a few of the basic features.  Here are some resources if you'd like to learn more:

+ Check out the "Help" tab on FlyBase
+ Visit [FlyBase TV](https://www.youtube.com/c/FlyBaseTV), the FlyBase YouTube channel
+ Look at the FlyBase "[New to Flies](https://wiki.flybase.org/wiki/FlyBase:New_to_Flies?)" wiki page
